\section{Related Work}
\label{sec:related_work}

\para{Lexinvariant language models and in-context deciphering.}
\citet{huang2023lexinvariant} showed that a transformer trained from scratch with per-sequence random Gaussian embeddings---effectively a new random permutation every sequence---can achieve near-normal language modeling given sufficient context ($\sim$512 tokens with character-level vocabularies). Their model performs implicit Bayesian deciphering, and probing reveals $>$99\% cipher key recovery accuracy. The critical difference from our work is that their model is \emph{trained} to handle arbitrary permutations; we study a \emph{pre-trained} model that has never seen permuted input. Our results show that the capacity for in-context deciphering requires explicit training and does not emerge from the architecture alone. Relatedly, \citet{hao2025cipher} demonstrated that pre-trained transformers can learn substitution ciphers given paired plaintext-ciphertext examples in context, but this requires explicit supervision pairs rather than raw ciphertext alone.

\para{Permutation equivariance and invariance in transformers.}
\citet{xu2024permutation} proved that vanilla transformers are equivariant to inter-token (row) permutations: $\text{Enc}(\mP_R \mZ) = \mP_R \, \text{Enc}(\mZ)$, where $\mP_R$ permutes the token ordering. This holds for both forward and backward passes in encoders. Critically, their result concerns permutations of \emph{token positions}, not \emph{token identities}: shuffling which position each token appears in preserves the residual stream up to the same shuffle. Our bijective transformation is fundamentally different---we change \emph{which embedding vector} enters the residual stream at each position, breaking the equivariance guarantee.

\para{Impossibility of permutation inversion.}
\citet{alur2025impossibility} proved that decoder-only, attention-only transformers of any depth cannot learn to invert permutations due to the causal attention mask preventing backward information flow. They showed two escape routes: removing the causal mask (enabling bidirectional attention) or padding with scratch tokens. Our empirical results strongly corroborate this impossibility: not only does GPT-2 fail to invert the permutation, it shows zero evidence of even partial adaptation---the permutation-adjusted prediction match is exactly 0\%.

\para{Residual stream analysis and dynamics.}
The residual stream---the main communication channel through a transformer's layers---has been studied as both a static representation space and a dynamical system. \citet{elhage2021mathematical} introduced the ``residual stream'' framing and showed how attention heads and MLPs read from and write to this shared stream. \citet{belrose2023tuned} introduced the tuned lens, training affine probes per layer to decode residual stream states into vocabulary distributions more reliably than the raw logit lens \citep{nostalgebraist2020logitlens}. \citet{fernando2025dynamics} treated the residual stream as a dynamical system evolving across layers, finding attractor-like dynamics and strong continuity of individual units. \citet{shai2024belief} showed that belief states from the theory of optimal prediction are linearly represented in the residual stream. \citet{lawson2024residual} trained multi-layer sparse autoencoders across all residual stream positions, finding that individual features activate at different layers for different tokens. Our work contributes to this line by revealing how residual stream dynamics change under a systematic input perturbation, exposing a two-phase structure that is invisible under normal operation.

\para{Substitution cipher models.}
A growing body of work investigates transformers' ability to handle substitution ciphers. \citet{isik2024interchangeable} proposed dual-part token embeddings with shared semantic and randomized distinguishing components to handle interchangeable tokens. \citet{alice2025cipher} introduced a specialized architecture for bijective cryptogram solving in a single forward pass. Unlike these approaches, which design architectures for cipher tasks, we examine how a standard pre-trained model's representations fail when confronted with an implicit cipher it was never trained to solve.
